---
title: "EES 3310/5310 Lab #1"
subtitle: "Exercises from Chapter 2"
author: "Kelsea Best"
date: "Lab: Monday Aug. 28, 2017. Due: Friday, Sept. 1 at 9:00 AM"
output:
  pdf_document: default
  html_document: default
  github_document: default
---

```{r setup, include=FALSE}
# This section sets up some options on knitr for processing the RMarkdown file
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)
```
```{r initialize, include=FALSE}
# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
library(tidyverse)
library(stringr)

source("scripts/format_md.R")
```
# Instructions:

* Clone a local copy of the file repository from github.
* In your local repository, answer the exercises in the template
  `Lab_01_Activities.Rmd`.
* When you are done, knit your `.Rmd` file into a PDF file.
* Use git to commit your changes (including the edits to `Lab_01_Activities.Rmd`
  and the new PDF file) to your local git repository.
* Push the changes from your local git repository to github.

    The last changes that you push before the due date (Friday 
    Sept. 1 at 9:00 am) will be graded.

I advise using git to commit changes frequently as you work and push those commits
to github.



# Exercises from Chapter 2

## Exercise 2.1

Consider exercise 1 in Chapter 2, on p. 17 of our textbook, 
_Global Warming: Understanding the Forecast_:

> A joule (J) is an amount of energy, and a watt (W) is a rate of using energy, 
> defined as 1 W = 1 J/s. How many Joules of energy are required to run a 100-W 
> light bulb for one day?

```{r ex_2_1_a}
seconds_per_hour = 60 * 60 # number of seconds in one hour
hours_per_day = 24 # number of seconds in one day
light_bulb_power = 100
joules_per_day = light_bulb_power * seconds_per_hour * hours_per_day

joules_per_day
```

It can be hard to read a long number like that, so we can use R's format command to add commas:

```{r formatting_example}
format_md(joules_per_day, comma = TRUE)
```

**Answer:** A `r light_bulb_power` watt light bulb burns `r light_bulb_power` Joules per second, 
so the light bulb burns `r light_bulb_power * seconds_per_hour` Joules per hour and
`r joules_per_day` Joules per day.

> Burning coal yields about 30 &times; 10^6^ J of energy per kilogram of coal burned. 
> Assuming that the coal power plant is 30% efficient,
> how much coal has to be burned to light that light bulb for one day?

```{r ex_2_1_b}
joules_heat_per_kg_coal = 30.E+6
efficiency = 0.30
joules_electricity_per_kg_coal = joules_heat_per_kg_coal * efficiency
kg_coal_per_joule_electricity = 1 / joules_electricity_per_kg_coal
kg_coal_per_day = joules_per_day * kg_coal_per_joule_electricity

kg_coal_per_day
```

**Answer:** 1 kilogram of coal produces `r format_md(joules_heat_per_kg_coal, digits = 1)` joules of heat, which is converted into `r format_md(joules_electricity_per_kg_coal, digits = 1)` joules of electricity.
We can turn this around and figure that to get 1 joule of electricity takes 1 / (`r format_md(joules_electricity_per_kg_coal, digits = 1)`) = `r format_md(kg_coal_per_joule_electricity, digits = 1)` kg of coal,
so to get `r format_md(joules_per_day, digits = 1)` joules of electricity to power the light bulb for one day takes `r format_md(kg_coal_per_day, digits = 2)` kg of coal.

## Exercise 2.2

> A gallon of gasoline carries with it about 1.3 &times; 10^8^ J of energy. 
> Given a price of $3 per callon, how many Joules can you get for a dollar?

```{r ex_2_2_a}
E <- 1.3 * 10^8 #J/ gallon gasoline 
Price <- 3 #$/ gallon

Joule_dollar_gasoline <- E / Price #Joules per dollar 
Joule_dollar_gasoline 

format_md(Joule_dollar_gasoline, comma = TRUE, digits =2)

```

**Answer:** You can get `r format_md(Joule_dollar_gasoline, digits =1)` Joules per dollar from gasoline.

> Electricity goes for about $0.05 per kilowatt hour. A kilowatt hour is just a 
> weird way to write Joules because a watt is a joule per second, and a kilowatt 
> hour is the number of Joules one would get from running 1000 W time one hour
> (3,600 seconds). In the form of electricity, how many Joules can you get for a 
> dollar?

```{r ex_2_2_b}
Cost_e <- 0.05 #cost $ per kilowatt hour
Cost_e_Joule <- 0.05/(1000*60*60)

Joules_electricity_per_dollar <- 1/Cost_e_Joule
Joules_electricity_per_dollar

```

**Answer:** You can get `r format_md(Joules_electricity_per_dollar, digits = 1)` Joules electricity for one dollar. 

> A standard cubic foot of natural gas carries with it about 1.1 &times; 10^6^ Joules
> of energy. You can get about 5 &times; 10^5^ British Thermal Units (BTUs) of gas
> for a dollar, and there are about 1,030 BTUs in a standard cubic foot. 
> How many Joules of energy in the form of natural gas can you get for a dollar?

```{r ex_2_2_c}
natural_gas <- 1.1*10^6 #Joules per standard cubic foot natural gas
BTU_per_dollar <- 5*10^5  #BTUs per dollar 
conversion <- 1030 #number to BTUs in a standard cubic foot

joules_natgas_per_dollar <- natural_gas * BTU_per_dollar * (1/conversion)  
joules_natgas_per_dollar

```

**Answer:** You can get `r format_md(joules_natgas_per_dollar, digits = 1)` Joules of energy from natural gas for a dollar. 

> A ton of coal holds about 3.2 &times; 10^10^ J of energy and costs about $40.
> How many Joules of energy in the form of coal can you get for a dollar?

```{r ex_2_2_d}
E_ton_coal <- 3.2*10^10 #Joules energy per ton coal 
cost_coal <- 40 #cost in $ of one ton coal 

E_per_dollar_coal <- E_ton_coal/ cost_coal

E_per_dollar_coal
```

**Answer:** You can get `r E_per_dollar_coal` Joules of energy from coal for a dollar. 

> Corn oil costs about $0.10 per fluid ounce wholesale. A fluid ounce carries
> about 240 dietary Calories (which a scientist would call kilocalories).
> A dietary Calorie is about 4200 J. How many Joules of energy in the form of
> corn oil can you get for a dollar?

```{r ex_2_2_e}
cost_corn <- 0.1 #cost $ per fluid ounce corn oil 
E_fluid_ounce <- 240 #E in calories per fluid ounce
Calorie_to_Joule <- 4200 #Joules per dietary calorie 

joules_per_dollar_corn <- E_fluid_ounce*Calorie_to_Joule*(1/cost_corn)

joules_per_dollar_corn
format_md(joules_per_dollar_corn, comma = TRUE)
```

**Answer:** You can get `r format_md(joules_per_dollar_corn, digits = 1)` Joules of energy from corn oil per dollar.

> Rank these five energy sources from cheap to expensive. 
> What is the range of prices?

**Answer:** 

Coal = `r format_md(E_per_dollar_coal, digits = 1)` Joules/$

Natural gas = `r format_md(joules_natgas_per_dollar, digits =1)` Joules/$

Electricity = `r Joules_electricity_per_dollar` Joules/$

Gasoline = `r format_md(Joule_dollar_gasoline, digits = 1)` Joules/$

Corn = `r format_md(joules_per_dollar_corn, digits =1)` Joules/$

Coal < Natural gas < Electricity < Gasoline < Corn 

The range of prices is `r format_md(1/E_per_dollar_coal, digits = 1)` dollars per Joule (coal) to `r format_md(1/joules_per_dollar_corn, digits = 1)` dollars per Joule (corn). 

## Exercise 2.3 (Grad students only)

> This is one of those job-interview questions to see how creative you are, 
> analogous to one I heard: "How many airplanes are over Chicago at any given time?" 
> You need to make stuff up to get an estimate and demonstrate your management 
> potential. 
> The question is: _What is the efficiency of energy production from growing corn?_
> 
> Assume that sunlight deposits 250 W/m^2^ of energy on a corn field, averaging 
> over the day-night cycle. 
> There are approximately 4,200 J per dietary Calorie. 
> How many Calories of energy are deposited on a square
> meter of field over the growing season? 
>
> (Note: the word "calorie" has two different meanings. Physicists and chemists, 
> use "calorie" (with a lower-case "c") to refer to a thermodynamic unit of heat, 
> but nutritionists use the word Calorie (with a capital 'C')
> to mean 1 kilocalorie (1000 thermodynamic calories), so when you see "Calories" 
> on a food label, it means kilocalories. To keep this exercise simple, I have 
> edited the textbook version so we only need to think in terms of dietary Calories.)


```{r ex_2_3_a}
Sun <- 250 #W/m^2
conversion <- 4200 #Joule/ Calorie
#Note, one W = 1 Joule/second 
growing_season <- 100 #assumed days of growing season
seconds <- 100*24*60*60 #seconds in growing season 
calories_total <- Sun*(1/conversion)*seconds 
calories_total
```

**Answer:** `r format_md(calories_total, digits =2)` Calories of energy are deposited from the sun on a square meter of field during a 100 day growing season. 

> Now guess how many ears of corn grow per square
> meter, and guess what the number of dietary Calories is that you get for eating 
> an ear of corn. Compare the sunlight energy with the corn energy to get the 
> efficiency.

```{r corn_efficiency}
calories_ear <- 600 #Calories per ear of corn 
corn_density <- 20 #ears of corn per square meter
corn_calories_for_eating <- calories_ear*corn_density
corn_calories_for_eating
efficiency <- corn_calories_for_eating / calories_total
efficiency

```

**Answer:** The efficiency of corn energy that I calculated is `r format_md(efficiency, digits =2)` or 2.3%.

## Exercise 2.4

> The Hoover Dam produces $2 \times 10^{9}$ W of electricity. It is composed of 
> $7 \times 10^{9}$ kg of concrete.
> Concrete requires 1 MJ of energy (1 megajoule, 1,000,000 Joules) to produce 
> per kilogram. 
> How much energy did it take to produce the dam? How long is the "energy payback 
> time" for the dam?

```{r ex_2_4_a}
P_hoover <- 2*10^9 #W electricity 
concrete <- 7*10^9 #kg concrete 
E_concrete <- 1*10^6 #Joules to produce one kilogram concrete

total_E <- concrete * E_concrete
total_E

#1 Watt = 1 Joule/second 
payback_time <- total_E / P_hoover #payback time in seconds
payback_days <- payback_time/(60*60*24)
payback_days
```

**Answer:** It took `r total_E` Joules of energy to produce the dam. They payback time is `r format_md(payback_days, digits =1)` days. 

> The area of Lake Mead, formed by Hoover Dam, is 247 mi^2^. Assuming 250 W/m^2^ 
> of sunlight falls on Lake Mead, how much energy could you produce if instead of 
> the lake you installed solar cells that were 12% efficient? 
> (1 mile is 1609 meters; how many square meters are in a square mile?)

```{r ex_2_4_b}
area_lake <- 247 #mi^2 
sunlight <- 250 #W/m^2
efficiency_solar <- 0.12 #12% 
conversion_miles_meters <- 1609^2 #number of squared meters in one squared mile
E_solar = efficiency_solar*area_lake*conversion_miles_meters*sunlight
E_solar

```

**Answer:** You could produce `r format_md(E_solar, digits = 2)` W of energy. 

## Exercise 2.5

> It takes approximately $2 \times 10^{9}$ J of energy to manufacture 1 m^2^ of 
> crystalline-silicon photovoltaic cell. (Actually, the number quoted was 
> 600 kilowatt hours. Can you figure out how to convert kilowatt hours into 
> Joules?) Assume that the solar cell is 12% efficient, and calculate how long it
> would take, given 250 W/m^2^ of sunlight, for the solar cell to repay the energy 
> it cost for its manufacture.

```{r ex_2_5}
E_photovoltaic <- 2*10^9 #Joules per m^2 of crystalline-silicon photovoltaic cell 
kilowatt_hours <- 600
kilowatt_hours_to_joules <- (kilowatt_hours*1000*60*60)
kilowatt_hours_to_joules

Sun = 250 #W/m^2

payback_solar <- kilowatt_hours_to_joules/(Sun*0.12) #joules / W yields seconds because 1 W = 1 J/s 
payback_solar #seconds

payback_solar_years <- payback_solar/(60*60*24*365)
payback_solar_years

```

**Answer:** It would take `r format_md(payback_solar_years, digits =2)` years to repay the energy for the manufacturing of a solar panel with an area of 1 $m^2$. 

## Exercise 2.7

> Infrared light has a wavelength of about 10 $\mu$m. What is its wave number in
> cm^-1^?

```{r ex_2_7_a}
wavelength_infrared <- 10 #micrometers 
wavelength_cm <- 10/10^4
wavenumber <- 1/wavelength_cm
wavenumber
```

**Answer:** The wavenumber is `r wavenumber` $cm^{-1}$. 

> Visible light has a wavelength of about 0.5 $\mu$m. What is its frequency in 
> Hz (cycles per second)?

```{r ex_2_7_b}
c <- 3*10^8 #speed of light m/s
wavelength <- 0.5*10^-6 #wavelength in meters
frequency <- c/ wavelength
frequency
```

**Answer:** The frequncy is `r frequency` Hz. 

> FM radio operates at a frequency of about 40 kHz. What is its wavelength?

```{r ex_2_7_c}
c <- 3*10^8 #speed of light m/s
frequency2 <- 40*10^3 #Hz
wavelength2 <- c/frequency2
wavelength2
```

**Answer:** The wavelength is `r wavelength2` m. 
