---
title: "VSA Climate Change Lab 01"
subtitle: "Exercises from Chapter 2"
author: "Your name"
date: "Add date"
output:
  pdf_document: default
  html_document: default
  github_document: default
---

```{r setup, include=FALSE}
# This section sets up some options on knitr for processing the RMarkdown file
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)
```
```{r initialize, include=FALSE}
# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
library(tidyverse)
library(stringr)

source("scripts/format_md.R")
```
# Instructions:

* Clone a local copy of the file repository from github.
* In your local repository, answer the exercises in the template
  `Lab1.Rmd`.
* When you are done, knit your `.Rmd` file into a PDF file.
* Use git to commit your changes (including the edits to `Lab1.Rmd`
  and the new PDF file) to your local git repository.
* Push the changes from your local git repository to github.


## Exercise 1


> A joule (J) is an amount of energy, and a watt (W) is a rate of using energy, 
> defined as 1 W = 1 J/s. How many Joules of energy are required to run a 100-W 
> light bulb for one day?

```{r ex_2_1_a}


```

It can be hard to read a long number like that, so we can use R's format command to add commas:

```{r formatting_example}
#format_md(, comma = TRUE)
```

**Answer:** 

> Burning coal yields about 30 &times; 10^6^ J of energy per kilogram of coal burned. 
> Assuming that the coal power plant is 30% efficient,
> how much coal has to be burned to light that light bulb for one day?

```{r ex_2_1_b}

```

**Answer:** 


## Exercise 2

> A gallon of gasoline carries with it about 1.3 &times; 10^8^ J of energy. 
> Given a price of $3 per callon, how many Joules can you get for a dollar?

```{r ex_2_2_a}


```

**Answer:** 

> Electricity goes for about $0.05 per kilowatt hour. A kilowatt hour is just a 
> weird way to write Joules because a watt is a joule per second, and a kilowatt 
> hour is the number of Joules one would get from running 1000 W time one hour
> (3,600 seconds). In the form of electricity, how many Joules can you get for a 
> dollar?

```{r ex_2_2_b}


```

**Answer:**  

> A standard cubic foot of natural gas carries with it about 1.1 &times; 10^6^ Joules
> of energy. You can get about 5 &times; 10^5^ British Thermal Units (BTUs) of gas
> for a dollar, and there are about 1,030 BTUs in a standard cubic foot. 
> How many Joules of energy in the form of natural gas can you get for a dollar?

```{r ex_2_2_c}


```

**Answer:** 

> A ton of coal holds about 3.2 &times; 10^10^ J of energy and costs about $40.
> How many Joules of energy in the form of coal can you get for a dollar?

```{r ex_2_2_d}

```

**Answer:** 

> Corn oil costs about $0.10 per fluid ounce wholesale. A fluid ounce carries
> about 240 dietary Calories (which a scientist would call kilocalories).
> A dietary Calorie is about 4200 J. How many Joules of energy in the form of
> corn oil can you get for a dollar?

```{r ex_2_2_e}


```

**Answer:** 

> Rank these five energy sources from cheap to expensive. 
> What is the range of prices?

**Answer:** 


